---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research: Peer assessment 1

## Loading and preprocessing the data

```{r, echo=FALSE}
## numbers >= 10^5 will be in scientific notation and rounded to 2 digits
options(scipen = 1, digits = 2)
```

```{r results="hide"}
unzip("activity.zip")
dati <- read.csv("activity.csv",sep=",")

dati$date   <- as.Date(dati$date)
dati$hour   <- dati$interval %/% 100
dati$minute <- dati$interval %%  100
dati$DateTime <- strptime(paste(dati$date, dati$hour, dati$minute), format="%Y-%m-%d %H %M")
```


```{r}
summary(dati)
```

## What is mean total number of steps taken per day?

First we calculate the total number of steps taken per day

```{r}
stepsaggr <- tapply(dati$steps, dati$date, sum, na.rm=TRUE)
```

We create the histogram of the total number of steps taken each day.

```{r fig.width=8, fig.height=7}
colors = c("red", "yellow", "green", "blue", "orange") 
hist(stepsaggr, main="Distribution of steps / day", col=colors, xlab="Steps")
```

Finally we calculate and report the mean and median of the total number of steps taken per day.

```{r}
sommario <- summary(stepsaggr)
```

Steps / day:
* mean: `r sommario[["Mean"]]`
* median: `r sommario[["Median"]]`

## What is the average daily activity pattern?

Time series plot of the 5-minute interval (x) and the average number of steps taken, averaged across all days (y).

```{r}
mediaintervalli <- tapply(dati$steps, dati$interval, mean, na.rm = TRUE)
disponibile <- require(lattice)
if (disponibile == FALSE) 
     {
      install.packages("lattice")
      library(lattice)
     }
xyplot(mediaintervalli ~ sort(unique(unclass(dati$interval))), type = "l", col = "red", xlab = "Interval", ylab = "Steps", main="Average daily activity pattern")
```

Determine the 5-minute interval with the highest average total steps.

```{r}
maxIntervallo <- mediaintervalli[abs(mediaintervalli - max(mediaintervalli)) < 1]
```

The interval with the maximum average number of steps is `r names(maxIntervallo)`.

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

```{r}
mancanti <-length((dati$steps[which(is.na(dati$steps)==TRUE)]))
```

The total number of missing values is `r mancanti`.

Strategy for filling in all of the missing values in the dataset: replaced missing values with the mean number of steps for each 5-minute interval. 

```{r}
intervalloMedia <-as.data.frame(tapply(dati$steps,dati$interval,mean,na.rm=TRUE))
names(intervalloMedia)<-c("steps")
```

New dataset that is equal to the original dataset but with the missing data filled in.

```{r}
dati[is.na(dati)] = intervalloMedia$steps
dati$steps<-round(dati$steps,0)
summary(dati)
```

Histogram of the total number of steps taken each day and mean and median total number of steps taken per day. 

```{r fig.width=8, fig.height=7}
colors = c("red", "yellow", "green", "blue", "orange") 
hist(tapply(dati$steps, dati$date,sum,na.rm=FALSE), col=colors,
     main="Distribution of steps taken each day",
     xlab="Steps")
meanNA<-mean(tapply(na.omit(dati$steps),dati$date,sum,na.rm=FALSE))
medianNA<-median(tapply(na.omit(dati$steps),dati$date,sum,na.rm=FALSE))
```

These values are higher than the estimates from the first part of the assignment. After changing the NA with the interval mean now the new mean is `r meanNA` and the new median is `r medianNA`.
Also the histogram is roughly the same.

# Are there differences in activity patterns between weekdays and weekends?

We use the dataset with the filled-in missing values.
We create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
Finally we make a panel plot containing a time series plot of the 5-minute interval (x) and the average number of steps taken, averaged across all weekday days or weekend days (y). 

```{r}
Sys.setlocale("LC_TIME", "English")
dati$date<-as.Date(dati$date)
dati$weekday<-weekdays(dati$date)
dati$weekday[dati$weekday=="Monday" | 
                 dati$weekday=="Tuesday" | 
                 dati$weekday=="Wednesday" | 
                 dati$weekday=="Thursday" | 
                 dati$weekday=="Friday" ]<-"weekday"
dati$weekday[dati$weekday=="Saturday" | dati$weekday=="Sunday"]<-"weekend"
dati$weekday<-factor(dati$weekday)
intAvgWeekday<-aggregate(steps ~ interval + weekday, data = dati, FUN= "mean" )

xyplot(intAvgWeekday$steps~intAvgWeekday$interval | intAvgWeekday$weekday, 
       type="l",
       layout=c(1,2),
       xlab="Interval", 
       ylab="Steps")
```


We can see that during the weekend people walks a little more than during weekdays.
